---
title: "Practice Cleaning Data"
output: html_document
date: "2025-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library (data.table)
install.packages("pROC")  # Only run this once
library(pROC)
install.packages("arrow")
library(arrow)
library(ggplot2)
```

# Here I am going to import my data, set my working directory, start subsetting the data

```{r}
# a good first step is to create a folder for your dataset and to store this markdown file in

setwd() # set working directory 

dataframe <- load.csv ("name of dataset.csv") # I would adapt this to the name of your crime dataset

compas_scores <- read.csv("compas-scores.csv")

View(dataframe) # check what your data looks like, what the names of the variables are

View(compas_scores)

#variables id dataset

str(compas_scores)

# transform the data to make race a dummy variable where Black == 1, white == 0

compas_scores$race_dummy <- ifelse(compas_scores$race == "African-American", 1,
                            ifelse(compas_scores$race == "Caucasian", 0, NA))

# this is the example code from lab 5, can you adjust it for the values you need?


# Create the new variable educated hispanics educated means educ_attain=>4
# sub_anes <- sub_anes %>% mutate(edu_hisp = ifelse(hisp == 1 & educ_attain >= 4,1,0))



# make my first subset data

subset1 <- dataframe[,c("variable 1", "variable 2", "variable 3")]

subset1 <- compas_scores[,c("v_decile_score","sex", "age", "race")]

# let's say you want to subset your data where a specific variable is equal to a specific outcome, like crime

subset2 <- dataframe[,c("variable 1", "variable 2", "variable 3", "c_charge_desc == battery")]

subset2 <- compas_scores[compas_scores$c_charge_desc == "Felony Petit Theft",
                         c("v_decile_score", "sex", "age", "race"),]


  
# now what if you want to subset your data to where a specific variable is equal to a few possible outcomes, use the & and function or the | or function to grab variables where the observation assumes one of a few particular values

subset3 <- dataframe[,c("variable 1", "variable 2", "variable 3", "c_charge_desc == battery" | "c_charge_desc == homocide"| "c_charge_desc == grand theft")]

subset3 <- compas_scores[,c("sex" + "race" + "age" + "priors_count" + "juv_fel_count", "c_charge_desc == Petit Theft" | "c_charge_desc == Possess Cannabis/20 Grams Or Less")]

# check how many obs for your subset data
nrow(subset1)
nrow (subset2)
summary (subset2$sex)

```

Once you have the subsets of the data you want, it's a good idea to note how
many observations you have overall, how many obs for the types of crimes you want, etc.

# Start Exploring the Data

```{r}
# check for incomplete observations, esp. for outcome

is.na(subset1$"outcome") # check your incomplete obs


#subset 2
subset2 <- subset2 [!is.na(subset2$"race"),]

#subset 3
subset3 <- subset3 [!is.na(subset3$race),]

## use = complete.obs argument, or similar na.rm = TRUE
subset1 <- subset1

summary(subset1, na.rm = TRUE) # gives summary w/o missing 

summary(subset1$race)

summary(subset2, na.rm = TRUE)

subset2$race <- as.factor(subset2$race)

summary(subset2$race)

```

I would also write down how many observations you have *after* you remove
the incomplete observations. This is the final amount of observations you will
be analyzing, which will be used for calculating your degrees of freedom (n-1).
```{r}
# Make race a categorical variable
compas_scores$race <- as.factor(compas_scores$race)
summary(compas_scores$race)

#Change reference level to Caucasian
compas_scores$race <- relevel(compas_scores$race, ref = "Caucasian")
```


```{r}

subset3 <- compas_scores %>%
  filter(race %in% c("African-American", "Caucasian"))

# 2. Create race dummy: 1 = Black, 0 = White
subset3$race_dummy <- ifelse(
  subset3$race == "African-American", 1,
  ifelse(subset3$race == "Caucasian", 0, NA)
)

# 3. Correlation between race (Black=1, White=0) and COMPAS violent decile score
cor(subset3$race_dummy, subset3$v_decile_score, use = "complete.obs")
```
```{r}
#Demographic Breakdowns of each Race
library(knitr)

# Create the frequency table and convert to data frame
race_table <- as.data.frame(table(compas_scores$race))

# Rename the columns
colnames(race_table) <- c("Race", "Count")

# Output table with new column names
kable(race_table, caption = "Race Distribution in COMPAS Dataset")

#Table Output
install.packages("kableExtra")
library(kableExtra)

#Save Table
kbl(race_table)

#Prop Table
race_prop <- as.data.frame(prop.table(compas_scores$race_dummy))

colnames(race_prop) <- c("Race", "Count")

kable(race_prop, caption = "Racial Breakdown in COMPAS")
```
```{r}

```


```{r}
#Gabby Examples
model_1 <- lm(raw score ~ sex + race + age, data = subset1 ) # for all crimes
summary(model_1) # this will give you summary statistics for your first model

 
model_2 <- lm(raw score ~ sex + race + age, data = subset1[,c_charg_desc == battery]) # this will give you a model where crimes are all battery
summary(model_2) # this will give you summary statistics for your second model

```


```{r}
#Base Model for all scores

model <- lm(v_decile_score ~ race + age + sex + priors_count + juv_fel_count, data = compas_scores)
summary(model)
```


```{r}
#Visuals for Base Model
library(ggplot2)
library(arm)
library(dotwhisker)


library(broom)
library(ggplot2)

#Coefficient Plot
model <- lm(v_decile_score ~ race + age + sex, data = compas_scores)
tidy_model <- tidy(model, conf.int = TRUE)

ggplot(tidy_model, aes(x = term, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  labs(title = "Model Coefficients", x = "Identity", y = "Estimate") +
  theme_classic() +
  coord_flip()
ggplot(tidy_model, aes(x = term, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Significance line
  labs(title = "Model Coefficients", x = "Identity", y = "Estimate") +
  theme_classic() +
  coord_flip()

```


```{r} 
#Base model (Grand Theft 3rd Degree)
model1 <- lm(v_decile_score ~ sex + race + age, data = subset1) 
summary(model1)

model2 <- lm(v_decile_score ~ sex + race + age + priors_count + juv_fel_count, 
             data = compas_scores[
               compas_scores$c_charge_desc == "Grand Theft in the 3rd Degree" &
               compas_scores$is_recid == 1, ])
summary(model2)

# Model for Grand Theft 3rd Degree
library(broom)
library(ggplot2)

tidy_model1 <- tidy(model1, conf.int = TRUE)
tidy_model1$model <- "Base Model"
tidy_model2 <- tidy(model2, conf.int = TRUE)
tidy_model2$model <- "Control Model"

coeffs <- rbind(tidy_model1, tidy_model2)

ggplot(coeffs[coeffs$term != "(Intercept)", ], aes(x = term, y = estimate, color = model)) +
  geom_point(position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, position = position_dodge(0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +  # <-- this line adds the reference for significance
  coord_flip() +
  labs(title = "COMPAS Scores (Grand Theft in the Third Degree)",
       y = "Coefficient Estimate (Risk Score)",
       x = "Predictor",
       caption = "Dashed line at zero indicates no statistical effect") +
  theme_minimal()
```
```{r}
# Base Model for Felony Petit Theft
model3 <- lm(v_decile_score ~ sex + race + age, data = subset2)

summary (model3)

# Full Model
model4 <- lm(v_decile_score ~ sex + race + age + priors_count + juv_fel_count, 
             data = compas_scores[
               compas_scores$c_charge_desc == "Felony Petit Theft" &
               compas_scores$is_recid == 1, ])
summary(model4)

# Visuals
library(broom)
library(ggplot2)

tidy_model3 <- tidy(model3, conf.int = TRUE)
tidy_model3$model <- "Base Model"
tidy_model4 <- tidy(model4, conf.int = TRUE)
tidy_model4$model <- "Control Model"

coeffs <- rbind(tidy_model3, tidy_model4)

ggplot(coeffs[coeffs$term != "(Intercept)", ], aes(x = term, y = estimate, color = model)) +
  geom_point(position = position_dodge(0.5)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, position = position_dodge(0.5)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +  # <-- this line adds the reference for significance
  coord_flip() +
  labs(title = "COMPAS Scores (Felony Petit Theft)",
       y = "Coefficient Estimate (Risk Score)",
       x = "Predictor",
       caption = "Dashed line at zero indicates no statistical effect") +
  theme_minimal()
```


```{r}
# Step 1: Start from your full COMPAS dataset
# Replace `compas` with the actual name of your dataset if different
compas_clean <- compas_scores %>%
  filter(
    race %in% c("African-American", "Caucasian"),
    is_recid %in% c(0, 1),
    !is.na(v_decile_score)
  )

# Step 2: Create race-specific subsets
black_clean <- compas_clean %>% filter(race == "African-American")
white_clean <- compas_clean %>% filter(race == "Caucasian")

# Step 3: Compute ROC objects
roc_black <- roc(
  response = black_clean$is_recid,
  predictor = black_clean$v_decile_score,
  levels = c(0, 1),
  direction = "<"
)

roc_white <- roc(
  response = white_clean$is_recid,
  predictor = white_clean$v_decile_score,
  levels = c(0, 1),
  direction = "<"
)

# Step 4: Get AUCs
auc_black <- auc(roc_black)
auc_white <- auc(roc_white)

auc_black
auc_white

# Step 5: Plot ROC curves with correct AUC labels
plot.roc(
  roc_black,
  col = "blue",
  main = "ROC Curves for COMPAS Violent Recidivism by Race",
  xlab = "False Positive Rate",
  ylab = "True Positive Rate"
)

lines.roc(roc_white, col = "red")

legend(
  "bottomright",
  legend = c(
    paste0("Black (AUC = ", round(auc_black, 2), ")"),
    paste0("White (AUC = ", round(auc_white, 2), ")")
  ),
  col = c("blue", "red"),
  lwd = 2
)
```

```{r}
#Total number of white and black defendants labeled high risk
# 1. Filter only Black and White defendants
high_risk_race_tbl <- compas_scores %>%
  filter(race %in% c("African-American", "Caucasian"))

# 2. Create a flag for "High Risk" (scores 7–10)
compas_clean <- high_risk_race_tbl %>%
  mutate(
    HighRisk = ifelse(v_decile_score >= 7, 1, 0),
    Race = ifelse(race == "African-American", "Black", "White")
  )

# 3. Count how many defendants in each race are labeled High Risk
highrisk_counts <- high_risk_race_tbl %>%
  group_by(race, v_score_text) %>%
  summarise(Total = n(), .groups = "drop") %>%
  mutate(PercentWithinRace = round(100 * Total / sum(Total), 1))

highrisk_counts
```


```{r}
# 1. Start from full dataset (replace `compas` with your actual dataset name)
compas_clean <- compas_scores %>%
  filter(
    race %in% c("African-American", "Caucasian"),
    is_recid %in% c(0, 1),
    !is.na(v_decile_score)
  ) %>%
  mutate(
    Race = ifelse(race == "African-American", "Black", "White")
  )

# 2. Focus on high-risk defendants only (8–10)
fig5_df <- compas_clean %>%
  mutate(
    ScoreHigh = v_decile_score >= 8,
    Recidivated = ifelse(is_recid == 1, "Yes", "No")
  ) %>%
  filter(ScoreHigh) %>%
  group_by(Race, Recidivated) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Race) %>%
  mutate(Percent = 100 * n / sum(n))

# 3. Filter for those who actually recidivated (Yes)
fig5_yes <- fig5_df %>%
  filter(Recidivated == "Yes")

# 4. Plot with colors and percent labels
ggplot(fig5_yes, aes(x = Race, y = Percent, fill = Race)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Percent, 1), "%")),
            vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Black" = "#1f78b4", "White" = "#33a02c")) +
  labs(
    title = "Recidivism Among High-Risk COMPAS Defendants by Race",
    x = "Race (High-Risk Defendants Only)",
    y = "Percent Who Recidivated"
  ) +
  ylim(0, 100) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )
```


# let's say you just want the coefficients and standard errors for your variables

sum1 <- summary(model2)
sum1$coefficients[,c(1,2)] # this will pull up just coeff and standard errors
```

We will do this more in lab with visualizations, but for your model you want
to report your coefficients, the significance of them (if any), and standard errors.

Then you want to interpret them narratively in context.

For example "In model 1, which controlled for age and sex, race was a statistically
significant predictor of risk score for those arrested on battery charges. The effect 
of being Black made the arrested person more likely to get a "High Risk" score by 
___ percentage points and more likely to get a "Moderate Risk" score by ___ percentage points."
