---
title: "risk-assessment-analysis"
author: "Max Taylor"
date: "2026-04-03"
output: html_document
---

#Install Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(dplyr)
library(readr)
library(ggplot2)
library(pROC)
library(broom)
library(knitr)
```

#Load and View Data
```{r}
# Update this path as needed
compas_scores <- read_csv("compas-scores.csv")

glimpse(compas_scores)
summary(compas_scores)
```

#Count of White and Black Defendants
```{r}
# Keep only Black and White defendants for the main comparison
compas_clean <- compas_scores %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  mutate(
    race = factor(race),
    race = relevel(race, ref = "Caucasian"),
    sex = factor(sex),
    race_dummy = ifelse(race == "African-American", 1, 0)
  )

summary(compas_clean$race)
table(compas_clean$race_dummy)
```

#Distribution of Black and White Defendants
```{r}
race_table <- compas_clean %>%
  count(race, name = "Count") %>%
  mutate(Percent = round(100 * Count / sum(Count), 1))

kable(race_table, caption = "Race Distribution in COMPAS Dataset")
```

#Correlation Score Between Race and Violent Risk Score
```{r}
compas_clean %>%
summarise(correlation = cor(race_dummy, v_decile_score, use = "complete.obs"))
```

#Baseline Model for all Races
```{r}
model_all <- lm(
  v_decile_score ~ race + age + sex + priors_count + juv_fel_count,
  data = compas_clean
)

summary(model_all)
```


#Coefficient Plot for Baseline Model
```{r}
tidy_all <- tidy(model_all, conf.int = TRUE)

ggplot(
  tidy_all %>% filter(term != "(Intercept)"),
  aes(x = term, y = estimate)
) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(
    title = "Model Coefficients",
    x = "Predictor",
    y = "Estimate"
  ) +
  theme_classic()
```

#Subset for Grand Theft in the Third Degree
```{r}
#Filter for Grand Theft in the Third Degree
grand_theft <- compas_clean %>%
  filter(c_charge_desc == "Grand Theft in the 3rd Degree")

nrow(grand_theft)
summary(grand_theft$race)

#Base Model for Grand Theft in the Third Degree 
model_gt_base <- grand_theft %>%
  lm (v_decile_score ~ sex + race + age,
  data = .)
  
#Full Model for Grand Theft in the Third Degree
model_gt_full <- grand_theft %>%
  lm(v_decile_score ~ sex + race + age + priors_count + juv_fel_count, data = .)

summary(model_gt_base)
summary(model_gt_full)

#Coefficient Plot for Grand Theft in the 3rd Degree Model
tidy_gt_base <- tidy(model_gt_base, conf.int = TRUE) %>%
  mutate(model = "Base Model")

tidy_gt_full <- tidy(model_gt_full, conf.int = TRUE) %>%
  mutate(model = "Full Model")

coeffs_gt <- bind_rows(tidy_gt_base, tidy_gt_full)

ggplot(
  coeffs_gt %>% filter(term != "(Intercept)"),
  aes(x = term, y = estimate, color = model)
) +
  geom_point(position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.2,
    position = position_dodge(0.5)
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    title = "COMPAS Scores: Grand Theft in the Third Degree",
    x = "Predictor",
    y = "Coefficient Estimate",
    caption = "Dashed line at zero indicates no statistical effect"
  ) +
  theme_minimal()
```

#Subset for Felony Petit Theft
```{r}
#Filter for Felony Petit Theft
petit_theft <- compas_clean %>%
  filter(c_charge_desc == "Felony Petit Theft")

nrow(petit_theft)
summary(petit_theft$sex)

#Model  for Felony Petit Theft
model_pt_base <- petit_theft %>%
lm(v_decile_score ~ sex + race + age, data = .)

model_pt_full <- petit_theft %>%
lm(v_decile_score ~ sex + race + age + priors_count + juv_fel_count, data = .)

summary(model_pt_base)
summary(model_pt_full)

#Coefficient Plot for Felony Petit Theft Model
tidy_pt_base <- tidy(model_pt_base, conf.int = TRUE) %>%
  mutate(model = "Base Model")

tidy_pt_full <- tidy(model_pt_full, conf.int = TRUE) %>%
  mutate(model = "Full Model")

coeffs_pt <- bind_rows(tidy_pt_base, tidy_pt_full)

ggplot(
  coeffs_pt %>% filter(term != "(Intercept)"),
  aes(x = term, y = estimate, color = model)
) +
  geom_point(position = position_dodge(0.5)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.2,
    position = position_dodge(0.5)
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    title = "COMPAS Scores: Felony Petit Theft",
    x = "Predictor",
    y = "Coefficient Estimate",
    caption = "Dashed line at zero indicates no statistical effect"
  ) +
  theme_minimal()
```

#ROC Curve by Race
```{r}
#Filter and Clean Data for Black and White Defendants
roc_data <- compas_clean %>%
  filter(is_recid %in% c(0, 1), !is.na(v_decile_score))

black_clean <- roc_data %>%
  filter(race == "African-American")

white_clean <- roc_data %>%
  filter(race == "Caucasian")

roc_black <- roc(
  response = black_clean$is_recid,
  predictor = black_clean$v_decile_score,
  levels = c(0, 1),
  direction = "<"
)

roc_white <- roc(
  response = white_clean$is_recid,
  predictor = white_clean$v_decile_score,
  levels = c(0, 1),
  direction = "<"
)

auc_black <- auc(roc_black)
auc_white <- auc(roc_white)

auc_black
auc_white

#ROC Curve Model
plot.roc(
  roc_black,
  col = "blue",
  main = "ROC Curve by Race",
  xlab = "False Positive Rate",
  ylab = "True Positive Rate"
)

lines.roc(roc_white, col = "red")

legend(
  "bottomright",
  legend = c(
    paste0("Black (AUC = ", round(auc_black, 2), ")"),
    paste0("White (AUC = ", round(auc_white, 2), ")")
  ),
  col = c("blue", "red"),
  lwd = 2
)
```

#Count of High Risk Classification Defendants
```{r}
highrisk_counts <- compas_clean %>%
  mutate(
    HighRisk = ifelse(v_decile_score >= 7, 1, 0),
    Race = ifelse(race == "African-American", "Black", "White")
  ) %>%
  group_by(Race, HighRisk) %>%
  summarise(Total = n(), .groups = "drop") %>%
  group_by(Race) %>%
  mutate(PercentWithinRace = round(100 * Total / sum(Total), 1))

kable(highrisk_counts, caption = "High-Risk Classification by Race")
```

#Recidivism Rates Among High-Risk Black and White Defendants
```{r}
#Filtering for Black and White Defendants Classified High Risk
fig5_df <- compas_clean %>%
  filter(is_recid %in% c(0, 1), !is.na(v_decile_score)) %>%
  mutate(
    Race = ifelse(race == "African-American", "Black", "White"),
    ScoreHigh = v_decile_score >= 8,
    Recidivated = ifelse(is_recid == 1, "Yes", "No")
  ) %>%
  filter(ScoreHigh) %>%
  group_by(Race, Recidivated) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Race) %>%
  mutate(Percent = 100 * n / sum(n))

#Filtering for High Risk Defendants who Recidivated
fig5_yes <- fig5_df %>%
  filter(Recidivated == "Yes")

#Bar Graph for Recidivism Rates for Black and White High Risk Defendants
ggplot(fig5_yes, aes(x = Race, y = Percent, fill = Race)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(
    aes(label = paste0(round(Percent, 1), "%")),
    vjust = -0.5,
    size = 5
  ) +
  labs(
    title = "Recidivism Among High-Risk COMPAS Defendants by Race",
    x = "Race (High-Risk Defendants Only)",
    y = "Percent Who Recidivated"
  ) +
  ylim(0, 100) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )

```